Table

### 2025-09-17 manuscript/terms_table.qmd
further updates

    Y = "This is the mathematical process of turning a condition variable into an indicator _sensu_ SEEA EA. It consists of three distinct steps: scaling, truncation, and transformation.

@@

    Normalisation is frequently referred to as **rescaling** in the SEEA EA literature, but we avoid  this terms because of the need to provide better differentiation between the three steps of the normalisation process, where scaling is one of them."

@@

    Y = "Spatial (sometimes called horizontal) aggregation is the process of summarising data from different sub-areas and taking this new value as the value for all constituent sub-areas. In means going from fine-resolution data to less fine or coarse resolution data. In related fields of study this operation may be described as _increasing the support_ or _reducing the grain_ [@bierkens2002]. Different summary functions can be used, including the area-weighted arithmetic mean, which is required under when aggregation indicator values in SEEA EA. The term _upscaling_ is sometimes used synonymously to our definition of spatial aggregation, but it also has a wider definition tied to the ambiguous use of the terms scale to mean either grain or extent. It's common to lump spatial aggregation together with some types of extra- or interpolation operation, such as averaging point data and attributing this value to entire polygons [@gruijter1990], or spatial estimation methods (e.g. geostatistics) and subsequent spatial sampling [@brus1997]. However, for the detailed deliberation in this study it is important to note the distinction between these."

@@

    X = "Inter- and extrapolation",
    Y = "Interpolation means to _fill in the gaps_, or to increase the density of the data. Extrapolation means to expand the spatial extent of the data or beyond the current data range. There is often no clear boundary between these two operations when applied in ECA as an area may for example be within the geographical range of the current data, but outside the environmental range. Many different methods are employed to do inter- or extrapolation within ECAs, such as statistical models (e.g. regression models) or geostatistical models (e.g. kriging). These methods are more or less sophisticaled ways to essentially _spread_ or learn from the available data to give values to areas where there is no data. This is distinct from the process and purpose of spatial aggregation, which seeks to simplify data."

@@

>>>>

    Y = "This is the mathematical process of turning a condition variable into an indicator _sensu_ SEEA EA. It consists of three distinct steps: scaling, truncation, and transformation (see also @fig-term).

@@

    Normalisation is frequently referred to as **rescaling** in the SEEA EA literature, but we avoid  this terms because of the need to provide better differentiation between the three steps of the normalisation process, where scaling is one of them. Also, in a parallel terminology, normalisation can be considered modeling, and spatial aggregation before and after normalisation can be referred to as pre or post processing, respectively [@vanbeurden1999]."

@@

    Y = "Spatial (sometimes called horizontal) aggregation is the process of summarising data from different sub-areas and taking this new value as the value for all constituent sub-areas. In means going from fine-resolution data to less fine or coarse resolution data. In related fields of study this operation may be described as _increasing the support_ or _reducing the grain_ [@bierkens2002]. Aggregation can be partial, resulting in two or more spatial areas with unique values, or total, if it results in just a single number summarising all the data. Different summary functions can be used, including the area-weighted arithmetic mean, which is required under when aggregation indicator values in SEEA EA. The term _upscaling_ is sometimes used synonymously to our definition of spatial aggregation, but it also has a wider definition tied to the ambiguous use of the terms scale to mean either grain or extent. It's common to lump spatial aggregation together with some types of extra- or interpolation operation, such as averaging point data and attributing this value to entire polygons [@gruijter1990], or spatial estimation methods (e.g. geostatistics) and subsequent spatial sampling [@brus1997]. However, for the detailed deliberation in this study it is important to note the distinction between these."

@@

    X = "Spatial prediction",
    Y = "It refers to any estimation of the value of a variable at an unobserved location, and therefore include both inter- and extraolation. **Interpolation** means to _fill in the gaps_, or to increase the density of the data. **Extrapolation** means to expand the spatial extent of the data or beyond the current data range. There is often no clear boundary between these two operations when applied in ECA as an area may for example be within the geographical range of the current data, but outside the environmental range. Many different methods are employed for spatial predictions within ECAs, such as statistical models (e.g. regression models) or geostatistical models (e.g. kriging). These methods are more or less sophisticaled ways to essentially _spread_ or learn from the available data to give values to areas where there is no data. This is distinct from the process and purpose of spatial aggregation, which seeks to simplify data."
    X = "Design-based and model based prediction",
    Y = "These are two fundamentally different ways of doing and thinking about spatialand statistical prediction. Design-based approaches relies on classical sampling theory and uses what is know about the sampling design to make spatial inferences. It is generally applied in more simple ways, such as with look-up tables, but model-assisted estimation using statistical models is also possible. Design-based approaches  produce things like the sample mean and sample variance and assumes the sampled values are correct (fixed), and uncertainty arises due to where we chose to sample (the locations are random). Model-based prediction uses geostatistical models, such as kriging and produces estimates such as the model mean or model variance. In this framework, the locations are considered fixed and the uncertainty arises due in the statistical model. Readers are referred to @brus2021 for more details."

@@

    X = "Post-stratfication",
    Y = "A statistical technique used to correct for a biased sample by forcing the sample to match the main population in some key characteristcis. In ECAs it is common to have data that lack a clear probability based sampling design, and therefore end up with more samples from certain strata (e.g. some geographic region or within some shared environmental condition). Post-stratification seeks to remove this unevenness in the _post_-processing, e.g. by doing a step-wise spatial aggregation via the same strata."
  )|>
